---
title: "Bulk RNA Seq Analysis Workflow"
author: "Patandeen Mishra"
date: "2026-05-10"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bulk RNA-Seq Analysis Workflow

This workflow demonstrates preprocessing, batch correction, PCA analysis, differential expression analysis, and pathway enrichment analysis using bulk RNA-Seq data.

---

## Load Libraries and Import Data

```{r}
library(tidyverse)

# Read metadata and raw count matrix
metadata <- read.csv("data/metadata.csv")
raw_counts <- read.csv("data/ileum_raw_counts.csv", check.names = FALSE, row.names = 1)

# Check dimensions
metadata
dim(metadata)
dim(raw_counts)

# Ensure unique gene names
rownames(raw_counts) <- make.unique(rownames(raw_counts))

# Keep only protein-coding genes
protein_coding_genes <- read_lines("data/pc_gene.txt")
raw_counts <- raw_counts[rownames(raw_counts) %in% protein_coding_genes, ]

# Match sample order between metadata and count matrix
metadata <- metadata[metadata$Sample %in% colnames(raw_counts), ]
metadata <- metadata[match(colnames(raw_counts), metadata$Sample), ]

# Confirm sample order
all(metadata$Sample == colnames(raw_counts))
```

---

## Handle Missing Values

```{r}
# Replace missing and non-finite values with zero
raw_counts[is.na(raw_counts)] <- 0
raw_counts[!is.finite(as.matrix(raw_counts))] <- 0

# Calculate percentage of missing genes per sample
sample_missing_pct <- colSums(raw_counts == 0) / nrow(raw_counts)

missing_df <- data.frame(
  Sample = colnames(raw_counts),
  Missing_Percentage = sample_missing_pct[colnames(raw_counts)]
)

# Save missing value statistics
write.csv(missing_df, "results/Percentage_of_missing_genes_per_samples.csv")

# Plot missing value distribution
png("figures/Histogram_percentage_of_missing_values_per_samples.png",
    width = 10, height = 8, units = 'in', res = 300)

hist(missing_df$Missing_Percentage,
     main = "Missing Genes per Sample",
     xlab = "Fraction Missing")

dev.off()

# Save cleaned raw matrix
write.csv(raw_counts, "results/raw_merged.csv")
```

---

## Batch Correction Using limma

```{r}
library(limma)

# Reorder metadata to match count matrix
metadata <- metadata[match(colnames(raw_counts), metadata$Sample), ]

# Confirm matching sample order
stopifnot(all(metadata$Sample == colnames(raw_counts)))

# Define batch and design matrix
metadata$Batch <- as.factor(metadata$Batch)
mod <- model.matrix(~ Condition, data = metadata)

# Perform batch correction
batch_corrected_counts <- removeBatchEffect(
  as.matrix(raw_counts),
  batch = metadata$Batch,
  design = mod
)
```

---

## Remove Unwanted Batches

```{r}
# Inspect batch distribution
print(table(metadata$Batch, metadata$Condition))

# Remove selected batches
exclude <- c("GSE183620", "GSE193677")

metadata$Batch <- droplevels(metadata$Batch)
metadata <- metadata[!metadata$Batch %in% exclude, ]
raw_counts <- raw_counts[, colnames(raw_counts) %in% metadata$Sample]

# Reorder metadata
metadata <- metadata[match(colnames(raw_counts), metadata$Sample), ]
metadata$Batch <- as.factor(metadata$Batch)

# Re-run batch correction
mod <- model.matrix(~ Condition, data = metadata)

batch_corrected_counts <- removeBatchEffect(
  as.matrix(raw_counts),
  batch = metadata$Batch,
  design = mod
)

# Save corrected matrix
write.csv(batch_corrected_counts, "results/limma_merged.csv")
```

---

## Principal Component Analysis (PCA)

```{r}
library(ggplot2)

# PCA before batch correction
raw_counts <- raw_counts[apply(raw_counts, 1, var) > 0, ]

pca_before <- prcomp(t(raw_counts), scale. = TRUE)

pca_before_df <- data.frame(
  pca_before$x,
  Condition = metadata$Condition,
  Batch = metadata$Batch
)

var_explained_before <- round(
  100 * (pca_before$sdev^2 / sum(pca_before$sdev^2)),
  2
)

# PCA after batch correction
batch_corrected_counts <- batch_corrected_counts[
  apply(batch_corrected_counts, 1, var) > 0,
]

pca_after <- prcomp(t(batch_corrected_counts), scale. = TRUE)

pca_after_df <- data.frame(
  pca_after$x,
  Condition = metadata$Condition,
  Batch = metadata$Batch
)

var_explained_after <- round(
  100 * (pca_after$sdev^2 / sum(pca_after$sdev^2)),
  2
)

# PCA plotting function by condition
plot_pca_condition <- function(pca_df, var_exp, title) {
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
    geom_point(size = 3, alpha = 0.8) +
    theme_minimal(base_size = 14) +
    labs(
      title = title,
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)")
    )
}

# PCA plotting function by batch
plot_pca_batch <- function(pca_df, var_exp, title) {
  ggplot(pca_df, aes(x = PC1, y = PC2, color = Batch)) +
    geom_point(size = 3, alpha = 0.8) +
    theme_minimal(base_size = 14) +
    labs(
      title = title,
      x = paste0("PC1 (", var_exp[1], "%)"),
      y = paste0("PC2 (", var_exp[2], "%)")
    )
}

# Generate PCA plots
p1 <- plot_pca_batch(
  pca_before_df,
  var_explained_before,
  "PCA Before Batch Correction"
)

p2 <- plot_pca_batch(
  pca_after_df,
  var_explained_after,
  "PCA After Batch Correction"
)

p3 <- plot_pca_condition(
  pca_before_df,
  var_explained_before,
  "PCA Before Correction - Condition"
)

p4 <- plot_pca_condition(
  pca_after_df,
  var_explained_after,
  "PCA After Correction - Condition"
)

# Save PCA plots
ggsave("figures/PCA_before_Batch.png", p1, width = 8, height = 6, dpi = 300)
ggsave("figures/PCA_after_Batch.png", p2, width = 8, height = 6, dpi = 300)
ggsave("figures/PCA_before_Condition.png", p3, width = 8, height = 6, dpi = 300)
ggsave("figures/PCA_after_Condition.png", p4, width = 8, height = 6, dpi = 300)
```

---

# Differential Expression Analysis

## Load Required Libraries

```{r}
library(limma)
library(edgeR)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)
library(tibble)
library(readr)
```

---

## Load Batch-Corrected Data

```{r}
# Read corrected count matrix
ileum_data <- read.csv("results/limma_merged.csv", row.names = 1)

# Match samples with metadata
ileum_data <- ileum_data[, colnames(ileum_data) %in% metadata$Sample]

# Check dimensions
dim(ileum_data)
```

---

## Create Design Matrix

```{r}
# Define experimental groups
group <- factor(metadata$Condition)

# Create design matrix
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)
```

---

## Differential Gene Expression Analysis

```{r}
# Fit linear model
fit <- lmFit(ileum_data, design)

# Define contrasts
contrast.matrix <- makeContrasts(
  DiseaseUntreated_vs_Healthy = UC_No - Healthy,
  levels = design
)

# Apply contrasts and empirical Bayes
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# Extract DEG results
top_genes <- topTable(fit2, adjust = "fdr", number = Inf)

# Save DEG table
write.csv(top_genes, "results/DEGsList.csv")

# Filter significant genes
DEGs <- top_genes[top_genes$adj.P.Val < 0.05, ]

# DEG summary statistics
total_DEGs <- nrow(DEGs)
upregulated <- sum(DEGs$logFC > 0)
downregulated <- sum(DEGs$logFC < 0)

cat(
  "At FDR < 0.05 - Total DEGs:",
  total_DEGs,
  "; Upregulated:",
  upregulated,
  "; Downregulated:",
  downregulated,
  "\n"
)
```

---

## PCA Visualization for DEG Analysis

```{r}
library(factoextra)

# Perform PCA on corrected data
pca_res <- prcomp(t(ileum_data), scale. = FALSE)

qq <- fviz_pca_ind(
  pca_res,
  label = "none",
  habillage = metadata$Condition,
  addEllipses = FALSE,
  mean.point = FALSE,
  title = "UC vs Healthy",
  pointsize = 3
) +
  theme_minimal(base_size = 15)

# Save PCA plot
png("figures/contrastPCA.png",
    width = 7,
    height = 5,
    units = 'in',
    res = 300)

plot(qq)
dev.off()
```

---

## Volcano Plot Visualization

```{r}
library(ggrepel)

# Genes of interest
highlightgenes <- c(
  "REG1A",
  "FDFT1",
  "SIK1",
  "DMBT1",
  "GGACT",
  "DUSP21"
)

# Assign significance categories
top_genes$Legend_Category <- "Not significant"

top_genes$Legend_Category[
  top_genes$adj.P.Val < 0.05
] <- "Significant"

top_genes$Legend_Category[
  rownames(top_genes) %in% highlightgenes &
    top_genes$adj.P.Val < 0.05
] <- "Reported in Papers"

# Define colors
legend_colors <- c(
  "Reported in Papers" = "red",
  "Significant" = "#1e90ff",
  "Not significant" = "grey"
)

# Label selected genes
label_df <- top_genes[
  rownames(top_genes) %in% highlightgenes &
    top_genes$adj.P.Val < 0.05,
]
# Reorder so "Highlighted in Paper" come last
top_genes$Legend_Category <- factor(top_genes$Legend_Category,
                                    levels = c("Not significant", "Significant", "Reported in Papers"))

# Generate volcano plot
volcano_plot <- ggplot(top_genes, aes(x = logFC, y = -log10(adj.P.Val),color = Legend_Category)) +
  geom_point(alpha = 0.7, size = 2.5) +
  scale_color_manual(values = legend_colors) +
  geom_vline(xintercept = c(0, 0), linetype = "dashed", color = "grey") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_text_repel(data = label_df,
                  aes(label = rownames(label_df)),
                  color = "black", size = 2.5,
                  box.padding = 0.4, point.padding = 0.3,
                  max.overlaps = 20, show.legend = FALSE) +
#  scale_x_continuous(limits = c(-3, 5)) +
  theme_minimal() +
  labs(title = "UC vs Healthy",
       x = "log2 Fold Change",
       y = "-log10 Adjusted P value",
       color = "Legend") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 11),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9)
  )

# Save volcano plot
ggsave(
  filename = "figures/VolcanoPlot.png",
  plot = volcano_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

---

## Reactome Pathway Enrichment Analysis

```{r}
# Prepare ranked gene list
res <- top_genes

# Convert gene symbols to ENTREZ IDs
ncbi_list <- clusterProfiler::bitr(
  rownames(res),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

res$names <- rownames(res)

res_mapped <- res %>%
  left_join(ncbi_list, by = c("names" = "SYMBOL")) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(ENTREZID, .keep_all = TRUE)  # remove duplicates

# Create ranked gene vector (logFC)
ngenes <- res_mapped$logFC
names(ngenes) <- res_mapped$ENTREZID
ngenes <- sort(ngenes, decreasing = TRUE)

# Run Reactome GSEA
enp_gsea <- gsePathway(
  geneList = ngenes,
  organism = "human",
  verbose = FALSE
)

# Convert IDs to readable gene names
enp_gsea <- setReadable(
  enp_gsea,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID"
)

# Extract pathway results
pathways <- enp_gsea@result %>%
  arrange(p.adjust)

# Select top pathways
top_pathways <- pathways %>%
  arrange(desc(abs(NES)))

# Plot Top 20 Reactome Pathways
top_pathways$Description <- str_wrap(
  top_pathways$Description,
  width = 50
)

top_pathways <- top_pathways[1:20, ] %>%
  mutate(Description = fct_reorder(Description, NES))

# Generate pathway enrichment plot
ggplot(top_pathways, aes(x = NES, y = Description, color = p.adjust, size = setSize)) +
  geom_point(alpha = 0.9) +
  scale_color_gradient(low = "#0072B2", high = "#D55E00", name = "FDR (p.adjust)") +
  scale_size(range = c(3, 10), name = "Gene Set Size") +
  labs(
    title = "Top 20 Reactome Pathways",
    x = "Normalized Enrichment Score",
    y = NULL
    ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 10, color = "black", face = "plain"),
    axis.text.x = element_text(size = 11),
    plot.title = element_text(size = 15, face = "plain", hjust = 0.5),
    plot.subtitle = element_text(size = 13, hjust = 0.5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0.5),
    legend.title = element_text(size = 12, face = "plain"),
    legend.text = element_text(size = 11)
  )

# Save pathway enrichment plot
ggsave(
  "figures/pathwaysplot.png",
  plot = pathway_plot,
  width = 8,
  height = 6
)

# Save pathway results
write.csv(pathways, "results/pathways.csv")
```

## DESeq2 Differential Expression Analysis

```{r}
library(DESeq2)

# Read metadata and raw count matrix
metadata <- read.csv("data/metadata.csv")
raw_counts <- read.csv(
  "data/ileum_raw_counts.csv",
  check.names = FALSE,
  row.names = 1
)

# Check dimensions
dim(metadata)
dim(raw_counts)

# Ensure unique gene names
rownames(raw_counts) <- make.unique(rownames(raw_counts))

# Keep only protein-coding genes
protein_coding_genes <- read_lines("data/pc_gene.txt")

raw_counts <- raw_counts[
  rownames(raw_counts) %in% protein_coding_genes,
]

# Remove confounded datasets
metadata <- metadata[
  !metadata$Batch %in% c("GSE183620", "GSE193677"),
]

# Keep matching samples only
raw_counts <- raw_counts[
  ,
  colnames(raw_counts) %in% metadata$Sample
]

# Reorder metadata
metadata <- metadata[
  match(colnames(raw_counts), metadata$Sample),
]

# Confirm sample order
all(metadata$Sample == colnames(raw_counts))

# Define condition factor
metadata$Condition <- factor(metadata$Condition)
metadata$Condition <- relevel(
  metadata$Condition,
  ref = "Healthy"
)

# Define batch factor
metadata$Batch <- factor(metadata$Batch)

# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(raw_counts),
  colData = metadata,
  design = ~ Batch + Condition
)

# Remove low-count genes
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Run DESeq2 analysis
dds <- DESeq(dds)

# Extract differential expression results
res <- results(
  dds,
  contrast = c("Condition", "UC_No", "Healthy")
)

# Order results by adjusted p-value
res <- res[order(res$padj), ]

# Convert to dataframe
res_df <- as.data.frame(res)

# Save DEG results
write.csv(
  res_df,
  "results/DESeq2_UC_vs_Healthy_results.csv"
)

# Count significant DEGs
upregulated_genes <- sum(
  res_df$log2FoldChange > 0 &
    res_df$padj < 0.05,
  na.rm = TRUE
)

downregulated_genes <- sum(
  res_df$log2FoldChange < 0 &
    res_df$padj < 0.05,
  na.rm = TRUE
)

# Print DEG summary
cat("Upregulated genes:", upregulated_genes, "\n")
cat("Downregulated genes:", downregulated_genes, "\n")
```
---
  
## Summary

This workflow demonstrates preprocessing, batch correction, PCA analysis, differential expression analysis, and pathway enrichment analysis for bulk RNA-Seq data using R.
